Using Fundamental Measure Theory to treat the Pair Distribution Function of the Inhomogeneous Hard-Sphere Fluid

### 2013-08-06 papers/pair-correlation/paper.tex
major edit of pair distribution paper Tested everything on bingley.

\documentclass[letterpaper,twocolumn,amsmath,amssymb,pre]{revtex4-1}
\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{color}
\usepackage{breqn}

\newcommand{\red}[1]{{\bf \color{red} #1}}
\newcommand{\blue}[1]{{\bf \color{blue} #1}}
\newcommand{\green}[1]{{\bf \color{green} #1}}
\newcommand{\rr}{\textbf{r}}
\newcommand{\refnote}{\red{[ref]}}

@@

% needsworklater is used to annotate bits that need work, but that we
% can postpone for a while.
\newcommand{\needsworklater}[1]{\emph{[#1]}}
% needsworknow is intended to prioritize stuff that needs fixing.
\newcommand{\needsworknow}[1]{\textcolor{red}{[\emph{#1}]}}

\begin{document}
\title{Using Fundamental Measure Theory to treat the two-point pair
  distribution function of the Inhomogeneous Hard-Sphere Fluid}

\author{Paho}
\author{???}
\author{David Roundy}
\affiliation{Department of Physics, Oregon State University, Corvallis, OR 97331}

%\pacs{61.20.Ne, 61.20.Gy, 61.20.Ja}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We develop and test an efficient approximation for the two-point
  pair distribution function of an inhomogeneous hard-sphere fluid.
\end{abstract}

\section{Why does pair distribution matter?}
\subsection{Interfaces are important}
\subsection{cDFT theory enables treatment of inhomogeneous systems}
\subsection{Pair distribution function is key input in thermodynamic perturbation theory}
The pair distribution function is a key input into thermodynamic
perturbation theory, which is widely used for treating dispersion
interactions in fluids.  In most cases, the hard-sphere fluid is used
as the reference system, making the pair distribution function of the
hard-sphere system an important quantity.
\subsection{Explain what the $a_1$ in thermodynamic perturbation theory is}

\section{What has been done before?}
\subsection{Gloor simple method, problematic}
Gil-Villegas\cite{gil1997statistical} approximates the first order inverse tenpurature Helmholtz Free energy term, which involves an integral of a potential multiplied by a radial distribution function $g^{HS}(r,\eta)$, by first taking the distribution function out of the integral and evaluating it at the correct mean distance $\zeta$, according to the mean value theorem.  They then approximate the pair correlation function $g^{HS}(\zeta,\eta)$ with the correlation at contact and at a effective $\eta_{eff}$.  The $\eta_{eff}$ is found by fitting a general function form to correct energy solutions $a_1$ with depenedent variable $\eta$ and

\subsection{Fischer accurate but slow}

\section{Triplet correlation function?}

@@

\begin{dmath}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, g_\sigma(\rr_1)) + g(|r_{12}|, g_\sigma(\rr_2))}{2}
\end{dmath}

@@

%% The homogeneoues limit $g_{sigm}$ can be calculated using the well
%% known and very successful Carnahan Starling equation while the
%% homogeneous limit radial distribution function can be calculated using
%% Monte-Carlo methods.  We used these calculations in a least squares
%% fit along the data points at increasing radial distances in order to
%% fit the perameters of our more general pair correlation function.  Our
%% function will therefore approach the most accurate solution possible
%% in the homogeneous limit.  Created a function that has the form of an
%% expansion in powers of $g_{sigma}$.  $g_{sigm}$ in the homogeneous
%% limit can be calculated using the well known and very successful
%% Carnahan Starling equation.  has a well known and Perameters in the
%% expansion where fit
\section{What does it look like (graphs that show the function itself)?}
\subsection{Fit to radial distribution function}
Figure~\ref{fig:radial-distribution} shows...
\begin{figure}
  \includegraphics[width=3cm]{figs/ghs-g}
  \includegraphics[width=3cm]{figs/ghs-g-ghs}
  \caption{Plot of the radial distribution function, with our fit.}\label{fig:radial-distribution}
\end{figure}
\subsection{Pair distribution function}
%% Figure~\ref{fig:pair-distribution} shows...
%% \begin{figure}
%%   %%\includegraphics[width=6cm]{figs/pair-correlation-2-395-0.pdf}
%%   \includegraphics[width=6cm]{figs/pair-correlation-2-395-5.pdf}
%%   %%\includegraphics[width=6cm]{figs/pair-correlation-2-395-10.pdf}
%%   \caption{Plot of the pair distribution function near a hard wall.
%%     In the interest of space and memeory, we should look at the function and decided which data to add?}\label{fig:pair-distribution}
%% \end{figure}

@@

Fu and Wu \cite{fu2005vapor} do I think the same as Gloor et al but maybe you (Dr. Roundy)
should take a look at it.  They reference Tang \cite{tang2008accurate} which I've
mentioned below in the Appendix.

@@

Fu and Wu \cite{fu2005vapor} do I think the same as Gloor et al but maybe you (Dr. Roundy)
should take a look at it.  They reference Tang \cite{tang2008accurate} which I've
mentioned below in the Appendix.

@@

\begin{dmath}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, \eta(g_\sigma(\rr_1))) + g(|r_{12}|, \eta(g_\sigma(\rr_2)))}{2}
\end{dmath}
Bad:
\begin{dmath}
  g^{(2)}(\rr_1,\rr_2) = g\left(|r_{12}|, \frac{\eta(g_\sigma(\rr_1)) + \eta(g_\sigma(\rr_2))}{2}\right)
\end{dmath}

The general idea is to write
\begin{dmath}
  g^{(2)}(\rr_1,\rr_2) = \Theta(r_{12}-\sigma)
  \left(1 + \sum_{\alpha} \frac{f_{\alpha}(g_\sigma(\rr_1)) + f_{\alpha}(g_\sigma(\rr_2))}{2}g_{\alpha}(r_{12})\right)\label{eq:gpair-approx}
\end{dmath}
where $f_{\alpha}$ and $g_{\alpha}$ are a set of functions that
satisfy the following constraint:
\begin{dmath}\label{eq:definealpha}
  \sum_{\alpha} f_{\alpha}(g_\sigma(\eta)) g_{\alpha}(r) \approx g(r; \eta)
\end{dmath}
which ensures that our pair distribution function reduces
(approximately) to the radial distribution function in the homogeneous
limit.  Equation~\ref{eq:gpair-approx} is constructed such that
Equation~\ref{eq:a1} can be evaluated in $O(N\log N)$ time, since all
the required convolutions may be performed with a fixed kernel.

@@

>>>>

\documentclass[letterpaper,twocolumn,amsmath,amssymb,pre]{revtex4-1}
\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{color}

\newcommand{\red}[1]{{\bf \color{red} #1}}
\newcommand{\rr}{\textbf{r}}
\newcommand{\refnote}{\red{[ref]}}

@@

\begin{document}
\title{Using Fundamental Measure Theory to treat the Pair
  Distribution Function of the Inhomogeneous Hard-Sphere Fluid}

\author{Paho Lurie-Gregg}
\author{Jeff B. Schulte}
\author{David Roundy}
\affiliation{Department of Physics, Oregon State University, Corvallis, OR 97331}

%\pacs{61.20.Ne, 61.20.Gy, 61.20.Ja}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  We have developed an approximation for the pair distribution
  function of hard spheres, which is based on the average value of the
  pair distribution function at contact described in
  Section~\ref{sec:contact}.  Our pair distribution is constructed so
  as to be computable with only single-site convolutions and to
  accurately reproduce the averaged distribution function at contact.
  This results in a functional that is efficient to compute, and
  predicts the pair distribution function as well as the most accurate
  existing analytical approximation.
\end{abstract}

\maketitle

\newcommand\saftlocaldft{}
% The following are papers that use a SAFT-based classical DFT with
% all the terms that should be non-local being non-local.
\newcommand\saftnonlocaldft{}

The standard approach in liquid state theory is to treat a liquid as a
hard-sphere reference fluid with attractive interactions that are
treated perturbatively~\cite{hansen2006theory}.  Recently there have
been considerable advances in the development of classical density
functionals \fixme{introduce DFT first} have been based on the
application of these theoretical approaches to inhomogeneous
densities~\cite{ jain2007modified, gloor2007prediction, gross2009density,
  %felipe2001examination, gloor2002saft,
  %gloor2004accurate, clark2006developing,
  kahl2008modified,
  % yu2002fmt-dft-inhomogeneous-associating,
  %fu2005vapor-liquid-dft, bryk2006density,
  hughes2013classical,
  %segura1998comparison, felipe2001examination,
  %gloor2002saft, gloor2004accurate, fu2005vapor-liquid-dft,
  bryk2006density, clark2006developing,
  kahl2008modified, gross2009density,
  %yu2002fmt-dft-inhomogeneous-associating,
  sundararaman2012computationally, marshall2013density}.  The
perturbation theory treatment of intermolecular interactions relies on
the pair distribution function of the reference fluid:
$g^{(2)}(\rr_1,\rr_2)$.  Unlike the radial distribution function of a
homogeneous fluid, there does not currently exist a tractable form for
the pair distribution function of an inhomogeneous hard-sphere fluid,
suitable for use in constructing a density
functional~\cite{gloor2007prediction, jain2007modified}.

\fixme{explain a bit of perturbation theory here?}

\section{Theoretical approaches}

The classic (and earliest) approach for computing the pair
distribution function is to use Percus' trick of treating one sphere
as an external field, and using the resultant equilibrium density to
find the pair distribution function~\cite{hansen2006theory}.  This
elegant approach lends itself to DFT, and can be used to compute and
plot the pair distribution function, but requires a full free-energy
minimization \emph{for each position} $\rr_1$ in
$g^{(2)}(\rr_1,\rr_2)$, and hence would be prohibitively expensive as
a tool in constructing a free energy functional.

The first canonical inhomogeneous configuration for the hard-sphere
fluid is the system consisting of a hard spheres at a hard wall.  In
1986, Plischke and Henderson solved the pair distribution function of
this system using integral equation theory under the Percus-Yevick
approximation~\cite{plischke1986pair}.  Ten years later,
G{\"o}tzelmann \emph{et al.} found the pair distribution function near
a hard wall using classical DFT by finding the direct correlation
function by taking a functional derivative of the free energy
functional and then solving the Ornstein-Zernicke
equation~\cite{gotzelmann1996structure}.  The approach of solving the
Ornstein-Zernicke equation for an inhomogeneous system is
computationally challenging~\cite{paul2003variational}.

A second inhomogeneous configuration that is of interest is the
test-particle configuration, in which one hard sphere is fixed in the
otherwise homogeneous fluid.  Where the hard-wall is a surface with no
curvature, the test-particle configuration has a surface with
curvature at the molecular length scale.  In this case, the density
gives the pair distribution function---this is just Percus'
trick---and the pair distribution function gives the triplet
distribution function.  The triplet distribution function was computed
by Gonz\'alez \emph{et al.} using the test-particle approach with
\emph{two} spheres fixed~\cite{gonzalez1999test}.

Lado recently introduced a new and improved efficient algorithm for
implementing integral equation theory for inhomogeneous fluids, which
computes $g^{(2)}(\rr_1,\rr_2)$~\cite{lado2009efficient}.  While this
approach is two orders of magnitude more efficient than previous
implementations, it remains a computationally expensive approach, and
unsuitable for repeated evaluation within a free-energy minimization.

\section{Analytic approximations}
There are few analytic approximations for the inhomogeneous pair
distribution function, which extend the radial distribution function
to inhomogeneous scenarios.  These works differ in what density they
use to evaluate $g(r)$.

Several recent works have cited a
mean-density approximation:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, \frac{\pi\sigma^3}{6}\tfrac12
  (n(\rr_1)+n(\rr_2))\right)
\end{align}
where $g(r,\eta)$ is the radial distribution function as a function of
radius and packing fraction $\eta$~\cite{gloor2007prediction,
  gross2009density}.  These works did not however use this approach,
which fails dramatically at higher densities, as the packing fraction
found from the mean density exceeds unity.

Approaches that do not diverge use an average of the density over some
volume. Fischer and Methfessel~\cite{fischer1980born} introduce the
approximation
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, n_3\left(\tfrac12
  (\rr_1+\rr_2)\right)\right)
  \label{eq:fischer}
\end{align}
where $n_3$ is one of the fundamental measures defined in Fundamental
Measure Theory~\cite{rosenfeld1989free}, which is an integral of the
density over a spherical volume:
\begin{align}
  n_3(\rr) = \int n(\rr')\Theta(\tfrac12 \sigma - |\rr-\rr'|) d\rr'
\end{align}
where $\sigma$ is the hard-sphere diameter.  Equation~\ref{eq:fischer}
is computationally awkward, because it relies on the midpoint
$\tfrac12(\rr_1+\rr_2)$.  For this reason Sokolowski and Fischer
modified this approach to use averages centered on the two points
$\rr_1$ and $\rr_2$:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12},
  \tfrac12(\bar{n}(\rr_1)+\bar{n}(\rr_2))\right)
  \label{eq:sokolowski}
\end{align}
where
\begin{align}
  \bar{n}(\rr) = \frac{3}{4\pi (0.8\sigma)^3}\int n(\rr')\Theta(0.8\sigma - |\rr-\rr'|) d\rr'
\end{align}
is the density averaged over a sphere with diameter
$0.8\sigma$~\cite{sokolowski1992role}.  The value 0.8 in this formula
was arrived at by fitting to Monte Carlo simulation.  Although
Eq.~\ref{eq:sokolowski} has the advantages of only involving density
averages at the points at which the pair distribution function is
desired, cannot be written as a single-site convolution, since the
convolution kernel depends on both points.

@@

\begin{align}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, g_\sigma(\rr_1)) + g(|r_{12}|, g_\sigma(\rr_2))}{2}
\end{align}

@@

%% The homogeneoues limit $g_{sigm}$ can be calculated using the well
%% known and very successful Carnahan Starling equation while the
%% homogeneous limit radial distribution function can be calculated using
%% Monte-Carlo methods.  We used these calculations in a least squares
%% fit along the data points at increasing radial distances in order to
%% fit the perameters of our more general pair correlation function.  Our
%% function will therefore approach the most accurate solution possible
%% in the homogeneous limit.  Created a function that has the form of an
%% expansion in powers of $g_{sigma}$.  $g_{sigm}$ in the homogeneous
%% limit can be calculated using the well known and very successful
%% Carnahan Starling equation.  has a well known and Perameters in the
%% expansion where fit
\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figs/ghs-g}\\
  \includegraphics[width=\columnwidth]{figs/ghs-g-ghs}
  \caption{Plot of the radial distribution function, with our fit.}\label{fig:radial-distribution}
\end{figure}
\section{What does it look like (graphs that show the function itself)?}
\subsection{Separable fit to the radial distribution function}
Figure~\ref{fig:radial-distribution} shows...

\subsection{Pair distribution function}
%% Figure~\ref{fig:pair-distribution} shows...
%% \begin{figure}
%%   %%\includegraphics[width=6cm]{figs/pair-correlation-2-395-0.pdf}
%%   \includegraphics[width=6cm]{figs/pair-correlation-2-395-5.pdf}
%%   %%\includegraphics[width=6cm]{figs/pair-correlation-2-395-10.pdf}
%%   \caption{Plot of the pair distribution function near a hard wall.
%%     In the interest of space and memeory, we should look at the function and decided which data to add?}\label{fig:pair-distribution}
%% \end{figure}

@@

Fu and Wu \cite{fu2005vapor} do I think the same as Gloor et al but
maybe you (Dr. Roundy) should take a look at it.  They reference Tang
\cite{tang2008accurate} which I've mentioned below in the Appendix.
\fixme{DJR: Not same as Gloor, they do a WDA approximation.}

@@

@@

\begin{align}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, \eta(g_\sigma(\rr_1))) + g(|r_{12}|, \eta(g_\sigma(\rr_2)))}{2}
\end{align}
Bad:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) = g\left(|r_{12}|, \frac{\eta(g_\sigma(\rr_1)) + \eta(g_\sigma(\rr_2))}{2}\right)
\end{align}

The general idea is to write
\begin{align}
  g^{(2)}(\rr_1,\rr_2) = \Theta(r_{12}-\sigma)
  \left(1 + \sum_{\alpha} \frac{f_{\alpha}(g_\sigma(\rr_1)) + f_{\alpha}(g_\sigma(\rr_2))}{2}g_{\alpha}(r_{12})\right)\label{eq:gpair-approx}
\end{align}
where $f_{\alpha}$ and $g_{\alpha}$ are a set of functions that
satisfy the following constraint:
\begin{align}\label{eq:definealpha}
  \sum_{\alpha} f_{\alpha}(g_\sigma(\eta)) g_{\alpha}(r) \approx g(r; \eta)
\end{align}
which ensures that our pair distribution function reduces
(approximately) to the radial distribution function in the homogeneous
limit.  Equation~\ref{eq:gpair-approx} is constructed such that
Equation~\ref{eq:a1} can be evaluated in $O(N\log N)$ time, since all
the required convolutions may be performed with a fixed kernel.

@@

%% G{\"o}tzelmann \emph{et al.} worked out the pair distribution function near a hard
%% wall by finding the direct correlation function by taking a functional
%% derivative of the free energy functional and then solving the
%% Ornstein-Zernicke equation~\cite{gotzelmann1996structure}.  They have plots like our pretty plots,
%% but with the fixed sphere one radius away from touching.
%% Interestingly, they plot the second bit as $z$, which enables them to
%% have just two axis values.  I think I prefer our ``total distance''
%% method, but there is a certain elegance to just plotting $z$ for the
%% curvy part as well as the straight-away part.  They found poor
%% quantitative agreement, particularly at contact.  They also studied
%% the pair correlation function in a hard slit~\cite{gotzelmann1997density}.

%% Integrating the Ornstein-Zernicke equation is challenging, requiring a
%% variational minimization itself~\cite{paul2003variational}.

%% Attard worked out the triplet correlation function, which is something
%% that we can access from the pair distribution function, if we consider
%% the pair correlation near a single hard sphere
%% solute\cite{attard1989spherically}.  Gonz\'alez \emph{et al.} wrote an
%% interesting paper using DFT (test-particle approach of Percus) and
%% Monte Carlo to compute the three-particle distribution function (same
%% as the triplet correlation function)\cite{gonzalez1999test}.  Their
%% theory compares favorably with the ``superposition approximation'',
%% and we could probably do something similar.

%% Plischke and Henderson worked out the pair correlation function near a
%% hard wall using Percus-Yevick, and compare with Monte Carlo
%% results\cite{plischke1986pair}.  They plot the pair correlation
%% function along interesting paths.  \fixme{Jeff read this paper for a
%%   while and does not understand what they do in oder to find the pair
%%   correlation function as compared to the radial distribution function}

%% Lado recently introduced a new and improved efficient algorithm for
%% implementing integral equation theory for inhomogeneous fluids, which
%% computes $g^{(2)}(\rr_1,\rr_2)$ among other
%% things\cite{lado2009efficient}.  Lado's work looks pretty awkward to
%% use, and their example is limited to 1D systems.  And the scaling is
%% still (so far as I can tell) $O(N^2)$, since their improvement is in
%% the prefactor, by developing a nice basis set.

%% Gloor \emph{et al} explain that little is known of the pair
%% distribution function of the hard-sphere fluid, and therefore use the
%% formula
%% \begin{align}
%%   g^{(2)}(\rr_1,\rr_2) \approx g\left(|\rr_1-\rr_2|, \frac{\rho(\rr_1)+\rho(\rr_2)}2\right)
%% \end{align}
%% They go on to further approximate the radial distribution function,
%% but we can just stop here.

%% Gloor et al \cite{gloor2007prediction} uses average density of two points:
%% \begin{align}
%% \overline{\rho} &\equiv \frac{\rho(\rr)+\rho(\rr')}{2}
%% \end{align}
%% and the uses bulk radial distribution for the reference system (hard
%% spheres) at this density, in order to approximate pair correlation.

%% \section{What has been done before?}
%% \subsection{Gloor simple method, problematic}
%% Gil-Villegas\cite{gil1997statistical} approximates the first order
%% inverse tenpurature Helmholtz Free energy term, which involves an
%% integral of a potential multiplied by a radial distribution function
%% $g^{HS}(r,\eta)$, by first taking the distribution function out of the
%% integral and evaluating it at the correct mean distance $\zeta$,
%% according to the mean value theorem.  They then approximate the pair
%% correlation function $g^{HS}(\zeta,\eta)$ with the correlation at
%% contact and at a effective $\eta_{eff}$.  The $\eta_{eff}$ is found by
%% fitting a general function form to correct energy solutions $a_1$ with
%% depenedent variable $\eta$ and

%% \subsection{Fischer accurate but slow}
